Do NBA Teams Avoid Trading within Their Own Division?

### 2022-05-10 Manuscript.Rmd
Back-implement in Rmd the changes jimi made to HTML

---
title: "Do NBA Teams Avoid Trading within Their Own Division?"
author: 
  - first_name: Jimi
    last_name: Adams
    orcid_id: "0000-0001-5043-1149"
    affiliation: University of Colorado
  - first_name: Michał
    last_name: Bojanowski
    orcid_id: "0000-0001-7503-852X"
    affiliation: Kozminski University
date: November 29, 2021
description: "Within US professional sports, trades within one's own division are often perceived to be disadvantageous. We ask how common this practice is. To examine this question, we construct a date-stamped network of all trades in the NBA between June 1976 and May 2019. We then use yearly weighted exponential random graph models to estimate the likelihood of teams avoiding within-division trade partners, and whether that pattern changes through time. In addition to the empirical question, this analysis serves to demonstrate the necessity and difficulty of constructing the proper baseline for statistical comparison. We find limited-to-no support for the popular perception."
output: 
  distill::distill_article:
    toc: true
    code_folding: false
link-citations: true
repository_url: https://github.com/mbojan/nba-trades
bibliography: nba_trades.bib
---

@@

<!-- Manually wrap in distill HTML tags to get a proper caption. -->
<div class="figure">
```{r animation, echo=FALSE}
htmltools::tags$iframe(src = 'NBA_Trades.html', width = '80%', height = '600px')
```
<p class="caption">Figure 3. **Dynamic Visualization of Annual Trade Networks** Note: Intra-division trades are highlighted in red. Node color and position refers to division.</p>
</div>

Our first step is to tabulate the number of trades observed both within and across divisions. This tabulation is presented in the form of what is referred to as a mixing matrix, which shows---at the level of divisions---how many trades occur between teams of the same division and between teams of differing divisions. We present these mixing matrices in two aggregations. The first represents all years prior to division expansion, when there were four divisions, a period which spans from 1976-2004. The second window began in 2005 and runs through the end of our analytic period at the end of the 2018-19 season, during which the teams were separated into six divisions. To provide a visual representation of the changes in these division-specific trade patterns across the examined period, Figure 3 presents a dynamic visualization of each year's trade network.^[The appendices also include full replication code for all analyzes conducted and the generation of this visualization.]

@@

It is worth mentioning that the models we have estimated in the results for Figure 5 are actually no different from what would have been possible through a generalized linear model predicting the (weighted) number of ties observed between pairs of teams. However, the value of introducing the weighted ERGM approach with these data is that if subsequent researchers want to take the next step to provide an explanatory account that further examines which teams are likely to trade with which others, this modeling framework would allow for the simultaneous estimation of predictive factors that variously operate at the node level (e.g., is a team that finished lower in the previous year's standings more likely to trade), the dyad level (e.g., are teams with higher total salaries more likely to trade with other teams who have lower total salaries), or structural features above the dyad (e.g., the tendency to form trade-triples or larger configurations through multi-team deals). The GLM framework would not be able to estimate each of these types of features in such a model without violating model assumptions (e.g., some of these are explicitly modeling the *dependency* between multiple trades). As such, we hope that the illustration if how the weighted ERGM could be beneficial serves as a useful introduction to this modeling framework, which could be extended by future researchers using the data we provide.

>>>>

---
title: "Do NBA Teams Avoid Trading within Their Own Division?"
author: 
  - first_name: jimi
    last_name: adams
    orcid_id: "0000-0001-5043-1149"
    affiliation: University of Colorado Denver, USA
  - first_name: Michał
    last_name: Bojanowski
    orcid_id: "0000-0001-7503-852X"
    affiliation: Kozminski University, Warsaw, Poland
date: "November 29, 2021"
description: "Within US professional sports, trades within one's own division are often perceived to be disadvantageous. We ask how common this practice is. To examine this question, we construct a date-stamped network of all trades in the NBA between June 1976 and May 2019. We then use yearly weighted exponential random graph models to estimate the likelihood of teams avoiding within-division trade partners, and whether that pattern changes through time. In addition to the empirical question, this analysis serves to demonstrate the necessity and difficulty of constructing the proper baseline for statistical comparison. We find limited-to-no support for the popular perception.<br/><br/>**Draft under review**."
output: 
  distill::distill_article:
    toc: true
    code_folding: false
    keep_md: true
link-citations: true
repository_url: https://github.com/mbojan/nba-trades
bibliography: nba_trades.bib
---

@@

<!-- Manually wrap in distill HTML tags to get a proper caption. -->
<div class="figure">
```{r animation, echo=FALSE}
htmltools::tags$iframe(src = 'NBA_Trades.html', width = '80%', height = '600px')
```
<p class="caption">Video 1. **Dynamic Visualization of Annual Trade Networks** Note: Intra-division trades are highlighted in red. Node color and position refers to division.</p>
</div>

Our first step is to tabulate the number of trades observed both within and across divisions. This tabulation is presented in the form of what is referred to as a mixing matrix, which shows---at the level of divisions---how many trades occur between teams of the same division and between teams of differing divisions. We present these mixing matrices in two aggregations. The first represents all years prior to division expansion, when there were four divisions, a period which spans from 1976-2004. The second window began in 2005 and runs through the end of our analytic period at the end of the 2018-19 season, during which the teams were separated into six divisions. To provide a visual representation of the changes in these division-specific trade patterns across the examined period, Video 1 presents a dynamic visualization of each year's trade network.^[The appendices also include full replication code for all analyzes conducted and the generation of this visualization.]

@@

It is worth mentioning that the models we have estimated in the results for Figure \@ref(fig:seasonal-ergms) are actually no different from what would have been possible through a generalized linear model predicting the (weighted) number of ties observed between pairs of teams. However, the value of introducing the weighted ERGM approach with these data is that if subsequent researchers want to take the next step to provide an explanatory account that further examines which teams are likely to trade with which others, this modeling framework would allow for the simultaneous estimation of predictive factors that variously operate at the node level (e.g., is a team that finished lower in the previous year's standings more likely to trade), the dyad level (e.g., are teams with higher total salaries more likely to trade with other teams who have lower total salaries), or structural features above the dyad (e.g., the tendency to form trade-triples or larger configurations through multi-team deals). The GLM framework would not be able to estimate each of these types of features in such a model without violating model assumptions (e.g., some of these are explicitly modeling the *dependency* between multiple trades). As such, we hope that the illustration if how the weighted ERGM could be beneficial serves as a useful introduction to this modeling framework, which could be extended by future researchers using the data we provide.
